---
title: "Exploring County Health with SHAP Values"
subtitle: "Understanding What Drives Poor Health Outcomes Across U.S. Counties"
author: "DS 400: Bayesian Statistics"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    df-print: paged
    fig-width: 10
    fig-height: 7
    embed-resources: true
    smooth-scroll: true
execute:
  warning: false
  message: false
  cache: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "100%"
)
```

## 🏥 Introduction: What Makes Communities Healthy?

What factors contribute most to poor health outcomes in U.S. counties? Is it education? Income? Access to healthcare? Environmental quality?

In this analysis, we'll use **County Health Rankings** data combined with **Random Forests** and **SHAP (SHapley Additive exPlanations)** values to uncover the key drivers of health outcomes across America.

::: {.callout-note icon="🎯"}
## Analysis Goals

-   Build a predictive model for poor health outcomes
-   Identify the most important features driving predictions
-   Understand how individual features affect health outcomes
-   Explore specific county predictions with SHAP values
:::

------------------------------------------------------------------------

## 📦 Load Required Packages

```{r}
#| label: load-packages
#| code-fold: false

library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(naniar)
library(randomForest)
library(treeshap)
```

::: callout-tip
## What are SHAP Values?

SHAP (SHapley Additive exPlanations) values explain the output of machine learning models by assigning each feature an importance value for a particular prediction. They're based on game theory and provide consistent, interpretable explanations!
:::

------------------------------------------------------------------------

## 📊 Load and Prepare the Data

### Import County Health Rankings Data

```{r}
#| label: load-data
#| code-fold: false

chr_data <- read_excel(here("mapping_machine/data", "2025_County_Health_Rankings_Data.xlsx"), sheet = 2, skip = 1)
chr_data <- chr_data %>%
  clean_names()
```

::: {.callout-tip icon="💡"}
## Challenge Problem #1

Create a map showing `percent_fair_or_poor_health` by county. You'll need to:

-   Use an AI assistant
    -   run glimpse(chr_data) in the console
    -   copy and paste the output into an AI
    -   prompt it to write simple and effective R code to make a choropleth map of `percent_fair_or_poor_health`by county

Give it a try!
:::

```{r}
# Load required libraries
library(tidyverse)
library(tigris)
library(sf)

# Use simple features (sf)
options(tigris_use_cache = TRUE)
options(tigris_class = "sf")

# Load county shapefiles (U.S. counties)
counties_sf <- counties(cb = TRUE, year = 2023)

# Prepare your data (make sure FIPS are 5-digit character codes)
chr_data <- chr_data %>%
  mutate(fips = str_pad(fips, width = 5, side = "left", pad = "0"))

# Join county geometries with your health data
map_data <- counties_sf %>%
  left_join(chr_data, by = c("GEOID" = "fips"))

# Plot choropleth map
ggplot(map_data) +
  geom_sf(aes(fill = percent_fair_or_poor_health), color = NA) +
  scale_fill_viridis_c(
    option = "plasma",
    na.value = "grey80",
    name = "% Fair or Poor Health"
  ) +
  labs(
    title = "Percent of Adults Reporting Fair or Poor Health by County",
    subtitle = "Source: County Health Rankings",
    caption = "Counties with missing data shown in grey"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )

#Prompt engineered by ChatGPT
```

------------------------------------------------------------------------

### Select Key Health Determinants

We're focusing on social, economic, and environmental factors that might influence health outcomes:

```{r}
#| label: select-variables
#| code-fold: false

chr_data <- chr_data %>%
  select(food_environment_index,
         percent_uninsured,
         percent_severe_housing_problems,
         percent_households_with_broadband_access,
         average_daily_pm2_5,
         percent_unemployed,
         income_ratio,
         percent_children_in_poverty,
         social_association_rate,
         preventable_hospitalization_rate,
         percent_completed_high_school,
         percent_some_college,
         percent_fair_or_poor_health)
```

::: {.callout-note collapse="true"}
## 📋 Variable Definitions

-   **Food Environment Index**: Measures access to healthy foods
-   **Percent Uninsured**: Population without health insurance
-   **Percent Severe Housing Problems**: Housing quality and affordability issues
-   **Broadband Access**: Households with internet access
-   **PM2.5**: Air pollution (particulate matter)
-   **Income Ratio**: Ratio of household income at 80th to 20th percentile
-   **Social Association Rate**: Membership in civic organizations
-   **Preventable Hospitalization Rate**: Hospital visits that could have been prevented
-   **Education Variables**: High school and college completion rates
-   **Percent Fair or Poor Health**: Our outcome variable (self-reported health status)
:::

------------------------------------------------------------------------

## 🔍 Exploratory Data Analysis

### Distribution of Our Outcome Variable

```{r, warning=FALSE, message=FALSE}
#| label: outcome-distribution
#| fig-cap: "Distribution of self-reported poor health across U.S. counties"

ggplot(chr_data, aes(x = percent_fair_or_poor_health)) +
  geom_histogram()
```

## 🕵️ Missing Data Investigation

Before building our model, let's check for missing values:

```{r}
#| label: missing-data
#| fig-cap: "Missing data patterns in our health dataset"

naniar::gg_miss_var(chr_data, show_pct = TRUE)
```

### Handle Missing Data

```{r}
#| label: drop-na
#| code-fold: false

chr_data <- chr_data %>%
  drop_na()
```

::: callout-warning
## Data Cleaning Note

We're using complete case analysis here (dropping rows with any missing values). In practice, you might want to explore more sophisticated imputation methods, especially if missingness is substantial or not random.
:::

------------------------------------------------------------------------

## 🌲 Build the Random Forest Model

Let's train a Random Forest to predict the percentage of people reporting fair or poor health:

```{r}
#| label: random-forest
#| code-fold: false

rf = randomForest(percent_fair_or_poor_health ~ ., data = chr_data, ntree = 500)
rf
```

::: {.callout-note icon="📈"}
## Model Performance

Look at the `% Var explained` metric. This tells us how much of the variance in health outcomes our model can explain using these social, economic, and environmental factors. Also look a the `Mean of squared residuals -`The average of the squared differences between the actual health outcomes and the model's predictions. Lower values indicate better predictive accuracy.
:::

------------------------------------------------------------------------

## 🔬 SHAP Analysis: Understanding Feature Importance

### Prepare for SHAP Calculation

```{r}
#| label: unify-model
#| code-fold: false

unified <- unify(rf, chr_data)
treeshap1 <- treeshap(unified,  chr_data[1:200, ], verbose = 0)
```

------------------------------------------------------------------------

## 📊 Feature Importance

Which factors matter most for predicting poor health outcomes?

```{r}
#| label: feature-importance
#| fig-cap: "Top 6 most important features for predicting poor health outcomes"
#| fig-height: 6

plot_feature_importance(treeshap1, max_vars = 6)
```

::: {.callout-important icon="🎯"}
## Key Insight

The features at the top have the biggest impact on predictions across all counties. These are the "leverage points" where interventions might have the greatest effect on community health!
:::

------------------------------------------------------------------------

## 📈 Feature Dependence: Deep Dive

### Education's Impact on Health

How does high school completion affect health outcomes?

```{r}
#| label: education-dependence
#| fig-cap: "Relationship between high school completion and SHAP values for health outcomes"

plot_feature_dependence(treeshap1, "percent_completed_high_school")
```

::: {.callout-note icon="💭"}
## Interpreting This Plot

-   **X-axis**: The actual values of the feature (% high school completion)
-   **Y-axis**: The SHAP value (how much this feature pushes predictions up or down)
-   **Pattern**: Look for trends - does higher education consistently lead to better health predictions?
:::

------------------------------------------------------------------------

::: {.callout-tip icon="🚀"}
## Challenge Problem #2

Create feature dependence plots for the next 4 most important variables from the feature importance plot above.

**Hint**: Use the same `plot_feature_dependence()` function but change the variable name.

``` r
# Template:
# plot_feature_dependence(treeshap1, "your_variable_name_here")
```

What patterns do you notice? Are the relationships linear or non-linear?
:::

```{r}
plot_feature_dependence(treeshap1, "percent_children_in_poverty")
```

```{r}
plot_feature_dependence(treeshap1, "food_environment_index")
```

```{r}
plot_feature_dependence(treeshap1, "percent_some_college")
```

```{r}
plot_feature_dependence(treeshap1, "percent_households_with_broadband_access")
```

------------------------------------------------------------------------

## 🎯 Individual County Analysis

Let's examine what drives the prediction for a specific county:

```{r}
#| label: individual-contribution
#| fig-cap: "SHAP contribution plot for observation 2: How each feature affects this county's prediction"

plot_contribution(treeshap1, obs = 2)
```

::: {.callout-note icon="🔍"}
## Reading the Contribution Plot

-   **Red bars**: Features pushing the prediction toward *worse* health outcomes
-   **Blue bars**: Features pushing the prediction toward *better* health outcomes\
-   **Length**: How strong the effect is

This shows you exactly which factors make this particular county's health outcomes better or worse than the baseline!
:::

------------------------------------------------------------------------

## 🎓 Reflection Questions

::: {.callout-warning icon="🤔"}
## Think About It

1.  **Policy Implications**: Based on the feature importance, which 2-3 areas should local governments prioritize for improving community health?

2.  **Causality vs Correlation**: Does this analysis prove that these features *cause* poor health outcomes? Why or why not?

3.  **Equity Considerations**: Do you notice any patterns that might indicate health disparities related to socioeconomic factors?

4.  **Model Limitations**: What important factors might be missing from this analysis? What would you add if you could?

5.  **Intervention Planning**: If you were a county health director, how would you use these SHAP values to design interventions?
:::

------------------------------------------------------------------------

## 💡 Key Takeaways

::: {.callout-tip icon="✨"}
## What We Learned

1.  **Random Forests** can effectively model complex health outcomes using multiple predictors
2.  **SHAP values** provide interpretable insights into both:
    -   Global feature importance (what matters overall)
    -   Local explanations (what matters for specific counties)
3.  **Health is multifaceted** - education, economics, environment, and social factors all play important roles
4.  **Machine learning + interpretability** tools like SHAP help us move from black-box predictions to actionable insights
:::

------------------------------------------------------------------------

## 📚 Additional Resources

-   [SHAP Documentation](https://shap.readthedocs.io/)
-   [County Health Rankings Methodology](https://www.countyhealthrankings.org/)
-   [treeshap R Package](https://github.com/ModelOriented/treeshap)
-   Random Forests in R: Breiman & Cutler's [Random Forest Page](https://www.stat.berkeley.edu/~breiman/RandomForests/)

------------------------------------------------------------------------

## 🔬 Your Turn: Extend the Analysis

::: {.callout-tip icon="🎯"}
## Extension Ideas

1.  **Try different sample sizes** for SHAP calculation (we used 200 counties)
2.  **Compare multiple observations** - pick 3-5 diverse counties and compare their contribution plots
3.  **Experiment with model parameters** - try different `ntree` values in the Random Forest
4.  **Temporal analysis** - if you have data from multiple years, track how feature importance changes over time
5.  **Subgroup analysis** - split by region or urbanicity and build separate models
:::

------------------------------------------------------------------------

*This analysis demonstrates the power of combining machine learning with interpretable AI techniques to understand complex public health outcomes. The insights generated can inform evidence-based policy decisions and targeted interventions!*

```{r}
#| label: session-info
#| echo: false
#| eval: false

# Uncomment to see package versions used
# sessionInfo()
```
